Embedded Scalars, Preferred Readings \& Intonation: {A}n Experimental Revisit

### 2014-09-09 text/paper/paper4.tex

\begin{table}
  \centering
  \begin{tabular}{lcccccccc@{\hskip 0.6cm}cccc}
    && \multicolumn{7}{c}{truth value at step}
    & \multicolumn{4}{c}{classification}
    \\ \cmidrule(l){3-9} \cmidrule(r){10-13}
    condition &
    & 1
    & 2
    & 3
    & 4
    & 5
    & 6
    & 7
    & \lit
    & \loc
    & \glb
    & err \\ \midrule
    \as-ntr 
    & T & 1 & 1 & \textbf{103} & 2 & \textbf{0} & 0 & 2 
    & 103 & 3 & 0 & 8 \\
    & F & 0 & 0 & 0 & 0 & 2 & 0 & \textbf{3} \\ 
    \addlinespace[0.25cm]
    \as-acc
    & T & 0 & 1 & \textbf{99} & 2 & \textbf{2} & 0 & 1 
    & 99 & 8 & 5 & 5 \\
    & F & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{8} \\
    \addlinespace[0.25cm]
    \es-ntr
    & T & 0 & 11 & 3 & 1 & \textbf{14} & - & - 
    & 78 & 14 & 0 & 22 \\
    & F & 0 & \textbf{0} & \textbf{78} & 5 & 2 & - & - \\
    \addlinespace[0.25cm]
    \es-acc
    & T & 0 & 10 & 3 & 0 & \textbf{16} & - & - 
    & 78 & 16 & 0 & 20 \\
    & F & 0 & \textbf{0} & \textbf{78} & 5 & 2 & - & -
  \end{tabular}
  \caption{Observed truth-value judgments in critical conditions. The
    left-hand part of the table shows the number of \emph{true} and
    \emph{false} responses obtained at each step in the sequence. The
    right-hand part shows the count for our imposed classification
    scheme. The answer counts indicative of relevant readings are
    given in bold-face on the left-hand side.}
  \label{tab:count_data_CC}
\end{table}

@@

Traditionalism easily explains the absence of global readings. Weak
traditionalism, introduced in Section~\ref{sec:traditionalism}, for
instance, would hold that the extra assumptions needed to derive
global readings (e.g., a speaker competence assumption) are not
available or only allow for epistemic weak implicatures that would
show as responses indicative of literal readings in our task. In
support of this view, traditionalists could argue that the task's
rather high processing demands lead to a suppression of global
readings, in line with the findings of
\citet{NeysDe-NeysSchaeken2007:When-People-Are} that working memory
load negatively affects the number of scalar implicature responses in
sentence verification tasks. The small number of local readings in
\as-sentences with contrastive stress is easily explained by adopting
the prosodic markedness hypothesis. What traditionalism does not
predict is the high number of answers indicative of local readings in
the \es-conditions. Resorting to the prosodic markedness hypothesis
does not help, because local readings were observed for accented as
well as for unaccented \es-sentences. It would also remain unclear why
prosodic cues should have such differential effects on \as- and
\es-sentences.

@@

Traditionalism, on the other hand, has to explain the possibility of
local readings of prosodically unmarked \es-sentences, and, perhaps,
also the large number of local readings in accented \es-sentences
compared to the much lower number in accented \as-sentences. To do so
within a strict traditionalist framework seems difficult. A first
options might be to account for local readings as the result of some
other phenomenon, such as typicality or contrast
\citep{Tielvan-Tiel2012:Embedded-Scalar,GeurtsTielvan-Tiel2013:Embedded-Scalar,Tielvan-Tiel2014:Quantity-Matter}.
Since this ties into a critical reflection on our design, we will
enlarge on the possibility of typicality- and contrast-based effects
in our setting. Eventually, we argue that contrast-effects alone are
unlikely to explain the observed high number of local answers in
\es-sentences, but that a ``modern traditionalist'' explanation of the
answer pattern might be available if we allow for the possibility that
\emph{exactly one} gets an unexpected ``referential interpretation.''

@@

A more promising alternative explanation for local readings of
\es-sentences is pursued by
\citet{GeurtsTielvan-Tiel2013:Embedded-Scalar} in response to the data
reported by \citet{ChemlaSpector2010:Experimental-Ev}. The main idea
is that visual contrast between an item that is connected to all, and
one that is connected to only some but not all of the relevant
elements may trigger exceptional local enrichment of \emph{some}. In
support of this idea,
\citeauthor{GeurtsTielvan-Tiel2013:Embedded-Scalar} show that the
strength of agreement to an \es-sentence on a 7-point Likert-scale
depends on how strong the relevant visual contrast is. When an item
with only some connections is presented alongside two universally
connected elements, the mean rate of agreement with a suitable
\es-sentence is significantly higher than when only one universally
connected element is present. So, maybe our \es-sequences similarly
provoked local readings because of an overemphasized visual contrast
between \emph{some but not all} and \emph{all}.

@@

On the other hand, there is a conceivable alternative explanation of
the puzzling data for \es-sentences that would also account for the
large and systematic error responses. Recall from
Section~\ref{sec:results} and Table~\ref{tab:count_data_CC} on
\pageref{tab:count_data_CC} that a surprisingly high number of
\emph{true} answers was given at the second step of the
\es-sequence. This answer type does not correspond to any of the three
candidate readings that the previous literature has focused on. As
suggested earlier, this answer type can be explained under the
assumption that \emph{exactly one} gets a reading similar to
\emph{there is (at least) one}, while the scalar item \emph{some}
would just receive its semantic interpretation. Such a reading of
\emph{exactly one}, although surprising, could actually be supported
by some theories of numerals and modifiers
\citep[e.g.][]{Geurts2006:Take-five,MartyChemla2014:Phantom-reading},
as the outcome of an ``existential closure'' type-shifting rule in the
sense of \citet{Partee1986:Noun-phrase-int} to the effect that
\emph{exactly one} is mapped onto an existential reading of the form
``there is a group with cardinality exactly one.''

What happens if \emph{some} is additionally strengthened in some way
or other under such a reading of \emph{exactly one}? If \emph{some} is
pragmatically strengthened under the scope of \emph{at least one} in a
pure traditionalist manner by adding the negation of a sentence type
\emph{at least one \dots all \dots}, we would expect \emph{true}
judgements at position 3 of the \es-sequence. Although exactly 3
\emph{true} answers occurred there for each neutral and accented
\es-conditions respectively, the main point to notice is that this
construal does not explain \emph{true} judgements at position 5. These
would therefore still seem indicative of local readings. However,
there is a slight variant of the hypothesized existentially closed
reading of \emph{exactly one}, that would explain \emph{true}
judgments at position 5 in an ``modern traditionalist'' manner. This
is a reading of \emph{exactly one} as ``there is a \emph{unique} group
with cardinality exactly one.'' Although this is admittedly only a
vague sketch of a possible line of explanation, it is not entirely
implausible that such a uniqueness requirement has participants look
for a distinguished referent, which they find on position 5. But if
the reading in question is about a unique discourse referent, we can
explain how the predicate \emph{is connected to some if its \dots} can
be pragmatically enriched in the desired fashion, along the lines of a
``modern traditionalist'' account that takes discourse effects into
account \citep[][Chapter~8.4]{Geurts2010:Quantity-Implic}. The
listener would then conclude that the referent in question was not
connected to all of its surrounding elements, because otherwise the
speaker would have attributed that to the now fixed referent in
question. This is parallel, so Geurts suggests, to the reasoning that
an utterance of ``Smith met a woman'' implicates that \emph{the woman
  referred to} was not Smith's mother and not that Smith didn't meet
his mother at all. But this alternative explanation is at present only
a tentative sketch of a possibility.

In conclusion, then, it appears that our data provides interesting
challenges to both traditionalism and grammaticalism.

>>>>

\begin{table}
  \centering
  \begin{tabular}{lcccccccc@{\hskip 0.6cm}cccc}
    && \multicolumn{7}{c}{truth value at step}
    & \multicolumn{4}{c}{classification}
    \\ \cmidrule(l){3-9} \cmidrule(r){10-13}
    condition &
    & 1
    & 2
    & 3
    & 4
    & 5
    & 6
    & 7
    & \lit
    & \loc
    & \glb
    & err \\ \midrule
    \as-ntr 
    & T & 1 & 1 & \textbf{103} & 2 & \textbf{0} & 0 & 2 
    & 103 & 3 & 0 & 8 \\
    & F & 0 & 0 & 0 & 0 & 2 & 0 & \textbf{3} \\ 
    \addlinespace[0.25cm]
    \as-acc
    & T & 0 & 1 & \textbf{99} & 2 & \textbf{2} & 0 & 1 
    & 99 & 8 & 2 & 5 \\
    & F & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{8} \\
    \addlinespace[0.25cm]
    \es-ntr
    & T & 0 & 11 & 3 & 1 & \textbf{14} & - & - 
    & 78 & 14 & 0 & 22 \\
    & F & 0 & \textbf{0} & \textbf{78} & 5 & 2 & - & - \\
    \addlinespace[0.25cm]
    \es-acc
    & T & 0 & 10 & 3 & 0 & \textbf{16} & - & - 
    & 78 & 16 & 0 & 20 \\
    & F & 0 & \textbf{0} & \textbf{78} & 5 & 2 & - & -
  \end{tabular}
  \caption{Observed truth-value judgments in critical conditions. The
    left-hand part of the table shows the number of \emph{true} and
    \emph{false} responses obtained at each step in the sequence. The
    right-hand part shows the count for our imposed classification
    scheme. The answer counts indicative of relevant readings are
    given in bold-face on the left-hand side.}
  \label{tab:count_data_CC}
\end{table}

@@

Traditionalism easily explains the absence of global readings. Weak
traditionalism, introduced in Section~\ref{sec:traditionalism}, for
instance, would hold that the extra assumptions needed to derive
global readings (e.g., a speaker competence assumption) are not
available or only allow for epistemically weak implicatures that would
show as responses indicative of literal readings in our task. In
support of this view, traditionalists could argue that the task's
rather high processing demands lead to a suppression of global
readings, in line with the findings of
\citet{NeysDe-NeysSchaeken2007:When-People-Are} that working memory
load negatively affects the number of scalar implicature responses in
sentence verification tasks. The small number of local readings in
\as-sentences with contrastive stress is easily explained by adopting
the prosodic markedness hypothesis. What traditionalism does not
predict is the high number of answers indicative of local readings in
the \es-conditions. Resorting to the prosodic markedness hypothesis
does not help, because local readings were observed for accented as
well as for unaccented \es-sentences. It would also remain unclear why
prosodic cues should have such differential effects on \as- and
\es-sentences.

@@

Traditionalism, on the other hand, has to explain the possibility of
local readings of prosodically unmarked \es-sentences, and, perhaps,
also the large number of local readings in accented \es-sentences
compared to the much lower number in accented \as-sentences. To do so
within a strict traditionalist framework seems difficult. A first
option might be to account for alleged local readings as the result of
some other phenomenon, such as typicality or contrast
\citep{Tielvan-Tiel2012:Embedded-Scalar,GeurtsTielvan-Tiel2013:Embedded-Scalar,Tielvan-Tiel2014:Quantity-Matter}.
Since this ties into a critical reflection on our design, we will
enlarge on the possibility of typicality- and contrast-based effects
in our setting. Eventually, we argue that contrast-effects alone are
unlikely to explain the observed high number of local answers in
\es-sentences, but that a ``modern traditionalist'' explanation of the
answer pattern might be available if we allow for the possibility that
\emph{exactly one} gets an unexpected ``referential interpretation.''

@@

A more promising alternative explanation for data indicative of local
readings of \es-sentences is pursued by
\citet{GeurtsTielvan-Tiel2013:Embedded-Scalar} in response to the data
reported by \citet{ChemlaSpector2010:Experimental-Ev}. The main idea
is that visual contrast between an item that is connected to all, and
one that is connected to only some but not all of the relevant
elements may trigger exceptional local enrichment of \emph{some}. In
support of this idea,
\citeauthor{GeurtsTielvan-Tiel2013:Embedded-Scalar} show that the
strength of agreement to an \es-sentence on a 7-point Likert-scale
depends on how strong the relevant visual contrast is. When an item
with only some connections is presented alongside two universally
connected elements, the mean rate of agreement with a suitable
\es-sentence is significantly higher than when only one universally
connected element is present. So, maybe our \es-sequences similarly
provoked local readings because of an overemphasized visual contrast
between \emph{some but not all} and \emph{all}.

@@

On the other hand, there is a conceivable alternative explanation of
the puzzling data for \es-sentences that would also account for the
large and systematic error responses. Recall from
Section~\ref{sec:results} and Table~\ref{tab:count_data_CC} on page
\pageref{tab:count_data_CC} that a surprisingly high number of
\emph{true} answers was given at the second step of the
\es-sequence. This answer type does not correspond to any of the three
candidate readings that the previous literature has focused on. As
suggested earlier, this answer type can be explained under the
assumption that \emph{exactly one} gets a reading similar to
\emph{there is (at least) one}, while the scalar item \emph{some}
would just receive its semantic interpretation. Such a reading of
\emph{exactly one}, although surprising, could actually be supported
by some theories of numerals and modifiers
\citep[e.g.][]{Geurts2006:Take-five,MartyChemla2014:Phantom-reading},
as the outcome of an ``existential closure'' type-shifting rule in the
sense of \citet{Partee1986:Noun-phrase-int} to the effect that
\emph{exactly one} is mapped onto an existential reading of the form
``there is a group with cardinality exactly one.''

What happens if \emph{some} is additionally strengthened in some way
or other under such a reading of \emph{exactly one}? If \emph{some} is
pragmatically strengthened under the scope of \emph{at least one} in a
pure traditionalist manner by adding the negation of a sentence type
\emph{at least one \dots all \dots}, we would expect \emph{true}
judgements at position 3 of the \es-sequence. Although exactly 3
\emph{true} answers occurred there for each neutral and accented
\es-conditions respectively, the main point to notice is that this
construal does not explain \emph{true} judgements at position 5, which
we classified as local responses. These would therefore still seem
indicative of local readings. However, there is a slight variant of
the hypothesized existentially closed reading of \emph{exactly one},
that would explain \emph{true} judgments at position 5. This is a
reading of \emph{exactly one} as ``there is a \emph{unique} group with
cardinality exactly one.''  Although this is admittedly only a vague
sketch of a possible line of explanation, it is not entirely
implausible that such a uniqueness requirement has participants, in a
first step of evaluation, look for a distinguished referent, which
they can find no sooner than on position 5 when the whole picture is
unravelled. The search for uniqueness would result in a witness for
the existential quantifier that appears to take a pragmatically
strengthened reading of \emph{some} into account, but that is not
necessarily a ``local reading'' in the sense of
grammaticalism. Moreover, although not strictly required to explain
the behavioral data, we can furthermore explain how the predicate
\emph{is connected to some if its \dots} can be pragmatically enriched
in the desired fashion, along the lines of a ``modern traditionalist''
account that makes use of interpretive mechanisms originally developed
to account for certain discourse phenomena
\citep[][Chapter~8.4]{Geurts2010:Quantity-Implic}. The listener would
then conclude that the referent in question was not connected to all
of its surrounding elements, because otherwise the speaker would have
attributed that to the now fixed referent in question. This is
parallel, so Geurts suggests, to the reasoning that an utterance of
``Smith met a woman'' implicates that \emph{the woman referred to} was
not Smith's mother and not that Smith didn't meet his mother at all.

To wrap up, we suggested amendements to grammaticalism and
traditionalism that could help either position explain the data we
observed. Whether these suggestions can be vindicated by further
empirical research is a matter that we must leave open. It remains,
however, that our data provides interesting challenges to both
traditionalism and grammaticalism.
